---
title: "Peer_Assignment_2"
author: "Zoheb Nensey"
date: "September 26, 2015"
output: html_document
---
# Insert Title Here 

## Synopsis 

## Data Processing
The data was downloaded from the course website as a csv.bz2 file, which can be read natively by the read.csv() function. 
```{r cache=TRUE}
storm_data <- read.csv("storm_data.csv.bz2")
```
The dplyr package was then loaded for grouping the data. 
```{r message=FALSE}
library(dplyr)
```

The data was then grouped by the event type and summed by both injuries and fatalities. 
A new variable (total_pop_health) was created summing the injuries and fatalities to explore total population damage by each type of weather. This dataset was then sorted by the total_pop_health variable. 

```{r}
storm_data_by_type <- storm_data %>%
      group_by(EVTYPE) %>%
      summarize(fatalities_sum=sum(FATALITIES),
                injury_sum=sum(INJURIES),
                total_pop_health = fatalities_sum+injury_sum)

storm_data_type_sorted <- storm_data_by_type[order(-storm_data_by_type$total_pop_health),]
```

## Results
```{r echo=FALSE}
biggest_event <- as.character(head(storm_data_type_sorted$EVTYPE, n=1))
event_fatalities <- as.character(head(storm_data_type_sorted$fatalities_sum,n=1))
event_injuries <- as.character(head(storm_data_type_sorted$injury_sum,n=1))
event_total <- as.character(head(storm_data_type_sorted$total_pop_health,n=1))
```
The worst type of weather event, as a category, was `r biggest_event`. It caused `r event_fatalities` fatalities and `r event_injuries` injuries for a total of `r event_total`.

```{r}
```